---
title: "MTH6139 Time Series"
subtitle: "Coursework 1 – Prophet Forecasting Analysis"
author: "Sarth Rahul Kavdikar"
date: "March 14, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---

# 1 Introduction

Time series forecasting is a common technique used in economics, business, and environmental science for prediction purposes. In this project, a classical time series dataset is used, and the Prophet time series forecasting system developed by Meta is employed for the purpose of prediction.

Prophet is a procedure for forecasting time series data with long-term trends and seasonalities that is simple to use.

In this project, the time series dataset used is the **AirPassengers dataset**.

---

# 2 Dataset Description

The dataset contains monthly passenger counts for international airline travel between 1949 and 1960.

```{r}
data(AirPassengers)

start(AirPassengers)
end(AirPassengers)
frequency(AirPassengers)
```

The series contains 144 observations with a frequency of 12, indicating monthly data.

---

# 3 Exploratory Analysis

We first visualise the time series to understand its structure.

```{r}
plot(AirPassengers,
     main="Monthly Airline Passenger Numbers",
     xlab="Year",
     ylab="Passengers (thousands)")
```

The plot reveals two important characteristics:

- A strong upward trend over time
- Increasing seasonal fluctuations

The seasonal pattern suggests that passenger numbers increase during certain months of the year.

---

# 4 Variance Stabilisation

The variability of the data increases as passenger numbers grow. To stabilise the variance we apply a logarithmic transformation.

```{r}
log_passengers <- log(AirPassengers)

plot(log_passengers,
     main="Log Transformed Passenger Data",
     xlab="Year",
     ylab="Log(Passengers)")
```

The transformation reduces the increasing spread of the data and helps produce a more stable time series.

---

# 5 Preparing Data for Prophet

Prophet requires a dataframe with two columns named **ds** (date) and **y** (value).

```{r}
library(prophet)
library(zoo)

time_values <- time(AirPassengers)

date_values <- as.Date(as.yearmon(time_values))

passenger_dataframe <- data.frame(
  ds = date_values,
  y = as.numeric(AirPassengers)
)

head(passenger_dataframe)
```

---

# 6 Prophet Forecasting Model

We now estimate a Prophet forecasting model using the passenger data.

```{r}
prophet_model <- prophet(passenger_dataframe)
```

The model automatically estimates trend and seasonal effects.

---

# 7 Forecasting Future Passenger Numbers

We generate predictions for the next 24 months.

```{r}
future_dates <- make_future_dataframe(prophet_model,
                                      periods = 24,
                                      freq = "month")

forecast_results <- predict(prophet_model, future_dates)

head(forecast_results)
```

---

# 8 Forecast Visualisation

```{r}
plot(prophet_model, forecast_results)
```

The plot displays the predicted passenger counts along with uncertainty intervals.

The forecast suggests continued growth in international airline travel.

---

# 9 Model Components

```{r}
prophet_plot_components(prophet_model, forecast_results)
```

The model separates the series into:

- Long-term trend
- Seasonal component

The seasonal component highlights recurring yearly travel patterns.

---

# 10 Conclusion

In this project, we worked with airline passenger data. The data shows strong growth trends that are characteristic of transportation demand. 

The data has been successfully modelled with the trend as well as the seasonality. The data is showing a trend that the number of passengers will continue to grow based on the current trends. 

This is a practical example of how we can use current forecasting tools to work with time-series data.

---

# References

Prophet Documentation  
https://facebook.github.io/prophet/

R Markdown Guide  
https://rmarkdown.rstudio.com